Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2013-11-06 response.tex

\documentclass[a4paper,10pt]{article}
%\documentclass[draftcls, onecolumn, 11pt]{IEEEtran}
%\documentclass[journal]{IEEEtran}

\usepackage{mathbf-abbrevs}
\input{defs}

%\usepackage{xr}
%\externaldocument{paper2}

\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}

\usepackage{amsmath,amsfonts,amssymb, amsthm, bm}

%\usepackage[square,comma,numbers,sort&compress]{natbib}

\usepackage{color}
\newcommand{\comment}[1]{\textcolor{red}{#1}}

\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\sinc}{\operatorname{sinc}}
\newcommand{\rect}[1]{\operatorname{rect}\left(#1\right)}

%opening
\title{Carrier phase and amplitude estimation for phase shift keying using pilots and data}
\author{Robby McKilliam, Andr\'{e} Pollok, Bill Cowley, I.\ Vaughan L.\ Clarkson and Barry Quinn  
%\thanks{}
}

\renewcommand{\theenumii}{(\alph{enumii})}
\renewcommand{\labelenumii}{\theenumii}

\begin{document}

\maketitle

\section*{Reviewer 1}

\begin{enumerate}
\item\textbf{Comment:} 
	This paper is concerned with carrier phase and amplitude estimation for PSK modulated signalling.
The considered least squares estimator seems to be only a minor extension of the one from [7]. Yet, more than 4 pages are dedicated to its derivation. Perhaps this section could be made more concise.
	\\\textbf{Response:}
	We have shortened Sec.\ II to make it more concise and improve readability.
\item\label{R1it2}\textbf{Comment:} 
	As stated by the authors themselves, the key results and main contributions of the paper involve the asymptotic properties of the least squares estimator. As far as I can see the derivation seems correct and complete, yet it is mathematically very involved and rather difficult to follow.
	\\\textbf{Response:}	
	For improved readability of the manuscript, we have now moved all of the lemmas into the appendix. We have also added a discussion of the implications and the practical relevance of the theorems to Sec.\ III (Sec.\ IV in the original manuscript).

	\comment{Stress practical relevance by showing ASRP PER results. Mention trials and patents. Point out that knowledge of the estimator statistics makes it possible to assess the estimator performance for a given number of symbols/pilot symbols at a certain SNR without having to run time consuming simulations. Also point out that the results are valid for a wider range of noise distributions, whilst CRBs need to be derived for a specific noise distribution\dots}
\item
A few minor remarks are:
\begin{enumerate}
\item\label{R1it3a}\textbf{Comment:} 
	Estimation from data symbols only is referred to as noncoherent, I rather consider a detector with non-data-aided synchronization as coherent (since it uses estimates of the synchronization parameters to detect the data).
	\\\textbf{Response:}
	We agree with the reviewer and apologise for causing confusion with the use of the terms \emph{coherent} and \emph{noncoherent} in the original manuscript. In the updated manuscript, we refer to an estimator that uses known pilot symbols as \emph{data-aided} and to one that relies on receiver decisions on the data symbols as \emph{decision-directed}.
\item\textbf{Comment:} 
	missing `,' after (5)
	\\\textbf{Response:} We thank the reviewer for spotting the missing comma and have added it.
\item\textbf{Comment:} 
	p.\ 19, at the bottom: sortindicIes (typo, delete 'i')
	\\\textbf{Response:} Thank you for pointing out this typographical error. We have corrected it.
\item\textbf{Comment:} 
	In Section VI, I would prefer a reference to the Lemma's in the Appendix at the start of the section.
	\\\textbf{Response:} We have added a reference to the Lemmas in the appendix at the beginning of the section. Please note that Sec.\ VI of the original manuscript is now Sec.\ III.
\item\textbf{Comment:} 
	Section VIII, ``Theorem 2 does not model this behaviour\dots'' can you please further explain why? + Comparison with the Cramer-Rao bound results from the existing literature would be interesting.
	\\\textbf{Response:}
	We have removed the aforementioned sentence from the manuscript as we feel it has caused more confusion than provided insight. We hope the following explanation clears up any confusion it may have caused.  Theorem 2 shows that for $L\rightarrow\infty$, $\sqrt{L}\sfracpart{\theta_0-\hat{\theta}}_\pi$ is normally distributed with zero mean and finite variance.  The support of $\sqrt{L}\sfracpart{\theta_0-\hat{\theta}}_\pi$ is the interval $[-\sqrt{L}\pi, \sqrt{L}\pi)$, i.e.\ $(-\infty, \infty)$ as $L\rightarrow\infty$.  For finite $L$, however, $\sqrt{L}\sfracpart{\theta_0-\hat{\theta}}_\pi$ has finite support and thus, cannot be exactly normally distributed.  Nevertheless, provided the variance of the noise is sufficiently small (i.e.\ the SNR sufficiently high), the distribution of $\sqrt{L}\sfracpart{\theta_0-\hat{\theta}}_\pi$ will be concentrated in a finite interval much smaller than $[-\sqrt{L}\pi, \sqrt{L}\pi)$ and is very close to a normal distribution.  On the other hand, as the SNR decreases, $\sqrt{L}\sfracpart{\theta_0-\hat{\theta}}_\pi$ approaches a uniform distribution on $[-\sqrt{L}\pi, \sqrt{L}\pi)$.  Consequently, the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$.  The argument for the case without pilots is analogous.

	\comment{Add CRBs to plots.} Please note that Sec.\ VIII of the original manuscript is now Sec.\ VI.
\end{enumerate}
\end{enumerate}

\section*{Reviewer 2}

\begin{enumerate}
\item
	General remarks
\begin{enumerate}
\item\textbf{Comment:} 
	Estimation of carrier phase and amplitude for $M$-ary PSK signals has been studied by the authors. Extending MackenthunÕs algorithm to pilots, a least square (LS) method is developed, where major parts of the manuscript are dedicated to the statistical analysis of the LS approach.

	Generally speaking it is to admit that both wording and writing follow the IEEE standards, although I would say that the list of References should be checked and updated in this respect.
	\\\textbf{Response:}
	We thank the reviewer for pointing us to the suggested references and have added \comment{some/most of} them. Specifically, we have included references [c, d, e]. \comment{Add the remaining ones?}
\end{enumerate}
\item
Some editorial comments
\begin{enumerate}
\item\textbf{Comment:} 
	Paragraph before (13): \dots can be computed from its predecessor \dots
	\\\textbf{Response:} Thank you for pointing out this typographical error. We have corrected it.
\item\textbf{Comment:} 
	Section III and VII: it would be helpful to provide $f(r, \phi)$
	\\\textbf{Response:}
	We thank the reviewer for pointing out the missing definition of $f(r, \phi)$ and have added it. Please note that in response to comment \ref{R4it2e} of reviewer 4, Sec.\ III has been slightly shortened and has been moved to the appendix.
\item\textbf{Comment:} 
	In the introductory section (paragraph before Eq.\ 3), the authors are talking about ``noncoherent detection where no pilot symbols exist''. In my mind this is somewhat confusing, because parameter estimation without knowing the underlying data is usually denoted as non-data aided (NDA) or decision-directed (DD), whereas when these data are known we have a data-aided (DA) situation [a]. On the other hand, coherent/noncoherent scenarios are given if the carrier phase is known/unknown to the receiver station.
	\\\textbf{Response:}
	Please refer to our response to comment \ref{R1it3a} of reviewer 1.
\end{enumerate}
\item
Major issues
\begin{enumerate}
\item\textbf{Comment:} 
	Figs.\ 1Ð3 illustrate the evolution of the phase MSE vs.\ SNR. Only in case that $|P| = 0$,
the LS results are compared to an alternative solution represented by the Viterbi and Viterbi (VV) algorithm. Why not doing the same with $|P| > 0$ by the following approach, which is definitely much simpler from the computational point of view (complexity in the order $O(L)$ instead of $O(L \log(L))$ required by LS) and the jitter performance will probably differ not very much from that of the LS variant (in Figs.\ 1Ð3 this can be seen for $|P| = 0$):
	\begin{itemize}
	\item
		If SNR $<$ SNRth, then the DA algorithm has to be used, else the VV algorithm must be applied.
	\item
		The SNRth is a threshold value known in advance, e.g., by the intersection of the Cramer-Rao lower bounds (CRLBs) for DA and NDA phase estimation [a, b].
	\item
		The SNR estimates might be obtained by a simple method based on second- and fourth-order moments [c].
	\item
		The same philosophy applies also to the estimation of amplitudes; for carrier- blind NDA results, the moment-based approach used for SNR estimation can be exploited.
	\end{itemize}
	\textbf{Response:}
	\comment{add response\dots}
\item\textbf{Comment:} 
	By detailed inspection of Fig.\ 4 showing the jitter variance for amplitude estimates, it is easily observed that $|P| < L$ performs better than $|P| = L$ in the medium-to-low SNR range. In my mind this is somewhat strange, since it should be the other way round (similar to the evolution of the phase curves in Figs.\ 1Ð3 and 5). Can you comment on this phenomenon ?
	\\\textbf{Response:}
	\comment{Replace variance plot by MSE plot.}
\item\textbf{Comment:} 
	In a much more powerful approach to the topic addressed in the current paper, the pilot sequence would be employed for a coarse estimation/correction of the carrier phase of the payload data. In the sequel, by embedding carrier phase recovery into an iterative (turbo) decoding scheme [d, e] or any other error correcting method [f, g], the jitter performance is close to the CRLB for DA estimation even at very small SNRs, which are mainly determined by the convergence region of the (turbo) decoder. By taking into account the fact that modern and powerful communication systems are not conceivable without any FEC scheme, it is not clear to me why I should choose your algorithm in such a context.
	\\\textbf{Response:}
	We have added a brief discussion of the need for sufficiently accurate coarse estimates to initialise an iterative code-aided synchronisation stage, citing [d, e].  We have also added a set of new simulation results, which show the packet error rate of an LDPC-coded satellite communication system to demonstrate the use and performance of the proposed estimator in a practical system.  \comment{Add details\dots}

	\comment{Add response related to [f, g]\dots}
\end{enumerate}
\end{enumerate}

\subsection*{Reviewer 2 references}
\begin{enumerate}\footnotesize
\renewcommand{\theenumi}{\alph{enumi}}
\renewcommand{\labelenumi}{[\theenumi]}
\addtolength{\itemsep}{-1.5ex}
\item
	Mengali, D'Andrea, \textit{Synchronization Techniques for Digital Receivers}, Plenum Press: New York, 1997.
\item
	Rice, Cowley, Moran, Rice, ``Cramer-Rao Lower Bounds for QAM Phase and Frequency Estimation'', \textit{IEEE Trans.\ Commun.}, Sept 2001.
\item
	Pauluzzi, Beaulieu, ``A Comparison of SNR Estimation Techniques for the AWGN Channel'', \textit{IEEE Trans.\ Commun.}, Oct.\ 2000.
\item
	Lottici, Luise, ``Embedding Carrier Phase Recovery into Iterative Decoding of Turbo-Coded Linear Modulations'', \textit{IEEE Trans.\ Commun.}, April 2004.
\item
	Noels et al., ``A Theoretical Framework for Soft-Information-Based Synchronization in Iterative (Turbo) Receivers'', \textit{Eurasip Journal Wireless Commun.\ and Networking}, April 2005.
\item
	D'Andrea, Mengali, Vitetta, ``Approximate ML Decoding of Coded PSK with no Explicit Carrier Phase Reference'', \textit{IEEE Trans.\ Commun.}, Feb./March/April 1994.
\item
	Vanelli, Salmi, Cioni, Corazza, Polydoros, ``A Performance Review of PSP for Joint Phase/Frequency and Data Estimation in Future Broadband Satellite Networks'', \textit{IEEE Journal on Selected Areas in Commun.}, Dec.\ 2001.
\end{enumerate}

\section*{Reviewer 3}

\begin{enumerate}
\item
	\textbf{Comment:} 
The paper studies the estimation of amplitude and phase due to the distortion introduced by the channel in communication systems with $M$-PSK signals. The estimator is derived from the observation of $L$ noisy samples taken at a rate of one sample per symbol. Part of the $L$ observed samples corresponds to pilot transmitted symbols, that are already known a priori by the receiver, while the rest of samples are associated with data symbols which are unknown at the receiver. The focus of the article is to determine the amplitude and phase without considering the estimation of the transmitted data like this problem is usually treated in the literature. The work modifies the Mackenthun algorithm to include samples corresponding to pilot symbols while the Mackenthun paper calculates a least squares estimator for a similar problem but without pilot symbols.

The major contribution of this paper is the description of the asymptotic properties of the least squares estimator for the phase and amplitude. Although it seems technically correct, the authors do not make clear the relevance of their work to signal processing applications and to timely problems. They limit themselves in presenting several lemmas and proofs. Therefore, I cannot recommend this manuscript for publication in the IEEE Transactions on Signal Processing.
	\\\textbf{Response:}
	We thank the reviewer for the comments. Please refer to our response to comment \ref{R1it2} of reviewer 1.
\end{enumerate}

\section*{Reviewer 4}

\begin{enumerate}
\item\textbf{Comment:}
	This work addresses the problem of carrier phase and amplitude estimation for $M$-ary PSK signals. Based on a modification of the Mackhentum's algorithm by inserting pilots, the main contribution of this manuscript is in fact the study of the asymptotic behavior of the least squares estimator for carrier phase and amplitude. Although the derivations seems correct, it is not clear to the reader why this method should be applicable in a communications context.  A cconsiderable part of the manuscript is dedicated to proofs of Lemmas and Theorems, which makes the paper rather difficult to read. There are several methods in the litterature that proposes the use of pilots to estimate the carrier phase followed by efficient decoding such as turbo decoding. How does the proposed method compare with the existing uncoherent methods that do not require estimation of the amplitudes? Even if the data are coded, what are the gains in terms of data recovery of the proposed method and traditional solutions? In my opinion, the authors fail to convincing on this aspect and this is an important point.
	\\\textbf{Response:}
	\comment{Stress practical relevance by showing ASRP PER results. In particular ST1, where the estimators are used to seed turbo sync. Mention trials and patents\dots}

	\comment{Not clear what the reviewer means by ``existing uncoherent methods that do not require estimation of the amplitudes''.}

\item
Some additional comments:
\begin{enumerate}
\item\textbf{Comment:} 
	The authors should better motivate the interest in using their proposed method in actual communication systems. The lack of motivation for addressing the problem raised in the manuscript is a weak aspect of this work.
	\\\textbf{Response:}
	\comment{Show PER ASRP results. Mention trials and patents\dots}
\item\textbf{Comment:} 
	The simulation results should bring comparisons with existing methods that use error correction and turbo decoding techniques, which have shown to perform reasonably well under adverse situations and low SNR levels.
	\\\textbf{Response:}
	\comment{Show PER ASRP ST1 turbo sync results. Comparison with other estimators?}
\item\textbf{Comment:} 
	In Fig.\ 4, the influence of $P$ on the performance is not clear to me and does not seem to be in accordance with other results in the manuscript.
	\\\textbf{Response:} 
	Please note that Fig.\ 4 in the original submission showed the variance of the amplitude estimator $\hat\rho$ (after bias removal), rather than the estimator MSE. This was done to validate the theoretical value of the asymptotic variance of $\hat\rho$ predicted by Theorem~2 against Monte-Carlo simulations. For consistency with the other figures and for easier interpretation, we have replaced the variance plot by a plot of the MSE of $\hat\rho$.
\item\textbf{Comment:} 
	It is important that the authors address an appropriate method for correcting the bias, as they mentioned in the text. A solution to provide the receiver with an estimate of $G(0)$ must be addressed in order to make this contribution applicable in a real communications context.
	\\\textbf{Response:} 
	Biased estimators are quite common in the communications literature and in fact, are often used in practice. Prime examples are state-of-the-art signal-to-noise ratio (SNR) estimators for $M$-PSK modulated signals such as those discussed by Pauluzzi and Beaulieu \cite{Pauluzzi2000}. With the exception of the data-aided SNR estimators, all of the estimators in \cite{Pauluzzi2000} are biased and particularly at low SNR, the bias can be large. To the best of the authors' knowledge, the value of the bias of these estimators has only been determined by computer simulation and has not been analysed theoretically. In contrast, the asymptotic analysis of the amplitude estimator $\hat\rho$ presented in Theorem~1 of our manuscript results in a closed-form expression for the bias and thus, lays the foundation for improving the estimator by addressing the bias. However, removal of the bias is beyond the scope of this manuscript.
\item\textbf{Comment:} \label{R4it2e}
	Some sectioms can be shortened to make the work more readable and concise. This is the case of Section II for instance. Section III is not necessary. It is too short and the main concepts are well known. A reference to a textbook would be sufficient. Section VII is also too short and is not justified.
	\\\textbf{Response:}
	We have shortened Sec.\ II to make it more concise and improve readability. Sec.\ VII has been removed from the revised manuscript.

	Furthermore, we agree with the reviewer that circularly symmetric random variables are well known, but would like to point out that Sec.\ III specifically discusses random variables of the form $Re^{j\varPhi}=1{+}W$, where $W$ is circularly symmetric. The joint pdf of $R$ and $\varPhi$ is a crucial ingredient for the asymptotic analysis presented in the manuscript and therefore, we feel the section should be retained. However, for improved readability, Sec.\ III has been slightly shortened and has been moved to the appendix.
\end{enumerate}
\end{enumerate}

\bibstyle{IEEEtran}
\bibliography{bib}

\end{document}

### 2013-11-14 paper.tex
moved paper back to old version and started modifying introduction.  Need to add that we've implemented it in asrp

In passband communication systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation (amplitude change).  These effects must be compensated for at the receiver. In this paper we assume that the time offset has been previously handled, and we focus on estimating the phase shift and attenuation.  We consider signalling constellations that have symbols evenly distributed on the complex unit circle such as binary phase shift keying (BPSK), quaternary phase shift keying (QPSK) and $M$-ary phase shift keying ($M$-PSK).  In this case, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M \geq 2$ is the size of the constellation.  We assume that some of the transmitted symbols are \emph{pilot symbols} known to the receiver and the remainder are information carrying \emph{data symbols} with phase that is unknown to the receiver.  So,
\[
s_i = \begin{cases}
p_i & i \in P \\
d_i & i \in D,
\end{cases}
\]
where $P$ is the set of indices describing the position of the pilot symbols $p_i$, and $D$ is a set of indices describing the position of the data symbols $d_i$.  The sets $P$ and $D$ are disjoint, i.e., $P \cap D = \emptyset$  where $\emptyset$ is the empty set, and we let $L = \abs{P \cup D}$ be the total number of symbols transmitted.

We assume that %time offset estimation has been performed and that 
$L$ noisy $M$-PSK symbols are observed by the receiver,  %The received signal after matched filtering is,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the receiver and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\begin{equation}\label{eq:SSdefn}
\begin{split}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - a p_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2,
\end{split}
\end{equation}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is additive white and Gaussian.  However, as we show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} where no pilot symbols exist ($P = \emptyset$).  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is the so called \emph{non-data aided}, sometimes also called \emph{non-decision directed}, estimator based on the paper of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the received signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{C}$ where $\angle$ denotes the complex argument and
\begin{equation}\label{eq:viterbiviterbi}
C = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
\end{equation}
Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not obvious how pilot symbols should be included. %Techniques are available but they are somewhat ad hoc~\cite{Cowley_ref_sym_carr_1998}.  
This problem does not occur with the least square estimator.

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Sweldens considered only the noncoherent setting, but we show in Section~\ref{sec:least-squar-estim}~that Mackenthun's algorithm can be modified to include pilot symbols. Our model includes the noncoherent case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

In the literature it has been common to assume that the data symbols $\{d_i, i \in D\}$ are of primary interest and that the complex amplitude $a_0$ is a nuisance parameter.  The metric of performance is correspondingly the \emph{symbol error rate}, or \emph{bit error rate}.  While estimating the data symbols is ultimately the goal, we take the opposite point of view here.  Our aim is to estimate $a_0$, and we treat the unknown data symbols as nuisance parameters.  This is motivated by the fact that in many modern communication systems the data symbols are \emph{coded}.  For this reason raw symbol error rate is not of interest at this stage.  Instead, we desire an accurate estimator $\hat{a}$ of $a_0$, so that the compensated received symbols $\hat{a}^{-1}y_i$ can be accurately modelled using an additive noise channel.  The additive noise channel is a common assumption for subsequent receiver operations, such as decoding.  The estimator $\hat{a}$ is also used in the computation of decoder metrics for modern decoders, and for interference cancellation in multiuser systems.  Consequently, our metric of performance will not be symbol or bit error rate, but $\sabs{\hat{a} - a_0}$. It will be informative to consider the carrier phase and amplitude estimators separately, that is, if $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number, then we consider $\sabs{\langle\theta_0 - \hat{\theta}\rangle_\pi}$ and $\sabs{\hat{\rho} - \rho_0}$.  The function $\fracpart{\cdot}_\pi$ denotes its argument taken `modulo $2\pi$' into the interval $[-\pi, \pi)$.  It will become apparent why $\langle\theta_0 - \hat{\theta}\rangle_\pi$ rather than $\theta_0 - \hat{\theta}$ is the appropriate measure of error for the phase parameter.  It is usual to write \emph{estimator} minus \emph{error}, i.e., $\hat{\theta} - \theta_0$ rather than $\theta_0 - \hat{\theta}$.  However, it is notationally convenient to use $\langle\theta_0 - \hat{\theta}\rangle_\pi$ rather than $\langle\hat{\theta} - \theta_0\rangle_\pi$ here.

>>>>

In passband communication systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation (amplitude change).  These effects must be compensated for at the receiver. In this paper we assume that the time offset has been previously handled and we focus on estimating the phase shift and attenuation.  We consider signalling constellations that have symbols evenly distributed on the complex unit circle such as binary phase shift keying (BPSK), quaternary phase shift keying (QPSK) and $M$-ary phase shift keying ($M$-PSK).  In this case, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M \geq 2$ is the size of the constellation.  We assume that some of the transmitted symbols are \emph{pilot symbols} known to the receiver and the remainder are information carrying \emph{data symbols} with phase that is unknown to the receiver.  So,
\[
s_i = \begin{cases}
p_i & i \in P \\
d_i & i \in D,
\end{cases}
\]
where $P$ is the set of indices describing the position of the pilot symbols $p_i$, and $D$ is a set of indices describing the position of the data symbols $d_i$.  The sets $P$ and $D$ are disjoint, i.e., $P \cap D = \emptyset$  where $\emptyset$ is the empty set, and we let $L = \abs{P \cup D}$ be the total number of symbols transmitted.

We assume that %time offset estimation has been performed and that 
$L$ noisy $M$-PSK symbols are observed by the receiver,  %The received signal after matched filtering is,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the receiver and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\begin{equation}\label{eq:SSdefn}
\begin{split}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - a p_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2,
\end{split}
\end{equation}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is white and Gaussian.  However, as we show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

% The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} (also know as \emph{blind}, \emph{non-data aided} or \emph{non-decision directed detection}) where no pilot symbols exist ($P = \emptyset$).  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is estimator of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the received signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{C}$ where $\angle$ denotes the complex argument and
% \begin{equation}\label{eq:viterbiviterbi}
% C = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
% \end{equation}
% Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not obvious how pilot symbols should be included. %Techniques are available but they are somewhat ad hoc~\cite{Cowley_ref_sym_carr_1998}.  
% This problem does not occur with the least square estimator.

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Sweldens considered what is called \emph{noncoherent detection} (also know as \emph{blind}, \emph{non-data aided} or \emph{non-decision directed detection}) where no pilot symbols exist ($P = \emptyset$).  We show in Section~\ref{sec:least-squar-estim}~that Mackenthun's algorithm can be modified to include pilot symbols. Our model includes the noncoherent case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

Aside from this extention of Mackenthun's algorithm, the primary purpose of this paper is to analyse the least square estimator of carrier amplitude and phase obtained by minimising~\eqref{eq:SSdefn}.  In the literature, analytic studies of non-data aided estimators of carrier phase have primarily involved the derivation of Cramer-Rao lower bounds under the assumption that the noise $\{w_i\}$ is complex Gaussian distributed~\cite{Delmas_exact_crb_psk_2008,Steendam_lowsnr_crb_2001,Cowley_crbs_phase_freq_1996,DAndrea_modified_bounds_1994}.  The bounds themselves are often complicated to compute and analyse and for this reason it is common to make approximations, known as \emph{modified} Cramer-Rao bounds~\cite{DAndrea_modified_bounds_1994}, or to analyse bound that are valid asymptoticially as the noise variance increases to infinity or decreases to zero~\cite{Moeneclaey_limit_crbs_1998,Steendam_lowsnr_crb_2001,Delmas_exact_crb_psk_2008}.  To the authors knowledge, all of the existing bounds are applicable only in the fully non-data aided case or in the completely data aided case, that is, when either no pilot symbols exist, or when no data symbols exist.  Deriving bounds for the common practical scenario where there is a mixture of pilot and data symbols might be possible, but would further complicate already complicated bounds.

For the case of carrier amplitude the situations is even more complicated.  A simulation study conducted by Pauluzzi~\cite{Pauluzzi2000} suggests that all known non-data aided estimators for carrier amplitude exhibit bias.  The bias is often small when the noise variance is small, but typically becomes significant as the noise variance increases.  Neverthess, Cramer-Rao bound have been derived for \emph{unbiased} estimators of carrier amplitude~\cite{Alagha_crb_snr_bpsk_qpsk_2001,Delmas_exact_crb_psk_2008}.  Unfortunately, these bounds may not be meaningful since known non-data aided amplitude estimators appear to be biased~\cite{Pauluzzi2000}.

In this paper we present an alternative approach to the study of estimators for carrier phase and amplitude.  Rather than compute bounds, we directly study the least squares estimators of carrier phase and amplitude asymptotically as the number of transmitted symbols $L\to \infty$.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent (asymptotically unbiased) and asymptotically normally distributed.  However, the least squares amplitude estimator is asymptotically biased and converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation $M$.  To our knowledge, this is the first time an analytic description of the bias or variance of non-data aided amplitude estimator has been obtained.  Our expressions for the asymptotic variance and bias are simple to compute.  %Our results are indirectly supported by the simulation study of Pauluzzi~\cite{Pauluzzi2000}.  We present our our set of simulations supporting our analytic results in Section~\ref{}.

%Two criticisms might be directed at our asymptotic approach when compared to the Cramer-Rao bounds already developed in the literature.  Perhaps the most obvious is that the results apply asymptotically as $L\to\infty$.  The simulation results we present in Section~\ref{} confirm that our asymptotic results accurately model the behaviour of the least squares estimator for modest numbers of symbols, for example $L=256$

One criticism that might be directed at our asymptotic approach when compared to the Cramer-Rao bounds already developed in the literature is that our specific results apply only to the least square estimator.  One response to this criticism is that, because the least squares estimator can be rapidly computed in $O(L\log L)$ operations and because it is the maximum likelihood estimator under the common assumption that the noise is white and Gaussian, then it is likely to be a popular practical choice.  Perhaps an even more convinving response is that, while the specific Theorems~\ref{thm:consistency} and~\ref{thm:normality} that we given apply only to the least square estimtor, the techniques that we develop to derive these results can likely be deployed for the analysis of other estimators.  For example, it would be reasonably straigthforward to extend our results to allow data and pilot symbols with varying constellation size, i.e. varying $M$.  For example, one might give more importance to certain data symbols and use BPSK ($M=2$) for these, but QPSK ($M=4$) for other less important symbols.  This is related to what is called \emph{unequal error protection} in the literature~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010} and is commonly used in practical systems.  Perhaps a more involved extension would be the inclusion of carrier frequency as a parameter to be estimated.  The techniques we present, combined with those from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}, would potentially allow this.
